Highly Available Transactions: Virtues and Limitations

### 2013-03-31 writeup/futurework.tex
joe comments

\textbf{Where do semantics-based concurrency control and weakened
  failure models lie in the HAT landscape?} The vast amount of
literature on semantics-based concurrency control provides hints as to
when coordination can be avoided while providing ``strongly
consistent'' outcomes for limited failure scenarios. We have performed
a narrow case study in this work, but a general taxonomization of
techniques such as escrow transactions~\cite{escrow} and
Sagas~\cite{sagas} could help answer the challenge of programming in a
weakly consistent environment. Additionally, HAT guarantees can be
strengthened by considering failure models with, say, bounded
asynchrony or service-level agreements from underlying
hardware. Additionally, while we have focused on systems that provide
a single availability mode, there are several interesting hybrids to
consider. For example, a system may provide ``best-effort'' stickiness
(or serializability!) in the absence of partitions, but, in the
presence of partitions, ``lose'' guarantees by having clients
reconnect to an available replica. This strategy provides stronger
guarantees in the absence of partitions but, to provide availability,
sacrifices them in the event of partitions.

>>>>

\textbf{Where do semantics-based concurrency control and weakened
  failure models lie in the HAT landscape?} The vast amount of
literature on semantics-based concurrency control provides hints as to
when coordination can be avoided while providing ``strongly
consistent'' outcomes for limited failure scenarios. We have performed
a narrow case study in this work, but a general taxonomization of
techniques such as escrow transactions~\cite{escrow},
Sagas~\cite{sagas}, and other long-running transaction
support~\cite{gray-virtues} could help answer the challenge of
programming in a weakly consistent environment. Additionally, HAT
guarantees can be strengthened by considering failure models with,
say, bounded asynchrony or service-level agreements from underlying
hardware. Additionally, while we have focused on systems that provide
a single availability mode, there are several interesting hybrids to
consider. For example, a system may provide ``best-effort'' stickiness
(or serializability!) in the absence of partitions, but, in the
presence of partitions, ``lose'' guarantees by having clients
reconnect to an available replica. This strategy provides stronger
guarantees in the absence of partitions but, to provide availability,
sacrifices them in the event of partitions.

### 2013-03-31 writeup/intro.tex
joe comments

Brewer's 2000 CAP Theorem stated that a highly available system cannot
provide strong consistency guarantees in the presence of network
partitions~\cite{gilbert-cap}. This result became a rallying call for
a generation of highly scalable but semantically weak system
designs. Most notably, the NoSQL movement embraced this approach,
offering production-use data platforms that eschewed strong
consistency guarantees but offered graceful scale-out to many
servers~\cite{bigtable, pnuts, dynamo}. The consistency concept
employed in the CAP Theorem came from the distributed systems
community and concerned reading the most recent version of a data item
that is replicated across servers.  However, the NoSQL movement
implicitly used this fairly narrow result to justify a lack of
general-purpose transactional functionality, or useful semantics for
groups of operations over multiple data items~\cite{hn,
  foundation-article}. Over a decade after CAP, we observe a dearth of
highly available systems providing transactional functionality.

Serializability, the gold standard of traditional ACID transactions,
is indeed not achievable given a requirement for high availability in
the presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-isolation}. Due
to their concurrency and overall performance benefits, weak isolation
models are widely employed, overwhelmingly as the default setting in
today's traditional ``ACID'' and up-and-coming ``NewSQL'' databases
and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alaternative semantics can be provided while
remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
has been poorly understood. We have a strong understanding of weak
isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-concurrency, spanner, granola,
  daudjee-session, krikellas-bargain, calvin} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. The distributed computing, programming language, and
parallel hardware literature also contains many consistency models for
single operations on replicated objects~\cite{pnuts, herlihy-art,
  cops, eiger, cac, sessionguarantees}. However, the literature has
largely ignored semantic guarantees for multiple operations operating
on multiple data items in a highly available distributed environment.

@@

Our investigation shows, somewhat surprisingly, that besides
serializability, only fairly strong models like Snapshot Isolation
(SI) and Repeatable Reads (RR) are not highly available, stemming from
their requirement to detect conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). In
contrast, Read Committed isolation, Transactional Atomicity, and many
of the weaker data consistency models from database and distributed
systems can be implemented in a highly available manner. For several
of these guarantees, such as Transactional Causal Predicate Cut
Isolation, we consider a modified form of high availiabity in which
clients ``stick to'' (i.e. have affinity with) at least one server (a
property which is implicit in the distributed systems literature but
which requires \textit{explicit} consideration in a client-server
replicated database context).

>>>>

The prior decade saw enormous churn in the design of widely-deployed
database backends, from the de facto use of transactional
RDBMSs~\cite{bernstein-concurrency, bernstein-book, gray-isolation,
  gray-virtues} to the widespread adoption of loosely-consistent
distributed key-value stores~\cite{bigtable, pnuts, dynamo}, with
increasing interest in returning to transactional
designs~\cite{spanner, walter, foundation-article, krikellas-bargain,
  eiger}. Instrumental in these trends was the 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions~\cite{brewer-slides}. For distributed systems
designers with requirements for ``always on,'' highly available and
low latency operation, CAP became an easily understandable indication
of the competing goals of semantic strength and availability. However,
as formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability, the ability to read the most recent write to a data
item that is replicated across servers. Yet this result---regarding a
fairly narrow form of distributed consistency---is often conflated
with ACID database properties~\cite{brewer-slides, hn,
  foundation-article}, leading to substantial confusion about the
relationships between replica consistency, transactional isolation, and high
availability.

Serializability, the gold standard of traditional ACID transactions,
is indeed not achievable given a requirement for high availability in
the presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-isolation,
  gray-virtues}. Due to their concurrency and overall performance
benefits, weak isolation models are widely employed, overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alaternative semantics can be provided while
remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
has not been well explored. We have a strong understanding of weak
isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-concurrency, spanner, granola,
  daudjee-session, krikellas-bargain, calvin} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. The distributed computing, programming language, and
parallel hardware literature also contains many consistency models for
single operations on replicated objects~\cite{pnuts, herlihy-art,
  cops, eiger, cac, sessionguarantees}. However, the literature has
largely ignored semantic guarantees for multiple operations operating
on multiple data items in a highly available distributed environment.

@@

Our investigation shows, somewhat surprisingly, that besides
serializability, only fairly strong models like Snapshot Isolation
(SI) and Repeatable Reads (RR) are not highly available, stemming from
their requirement to detect conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). In
contrast, Read Committed isolation, Transactional Atomicity, and many
of the weaker data consistency models from database and distributed
systems can be implemented in a highly available manner. For several
of these guarantees, such as Transactional Causal Predicate Cut
Isolation, we consider a modified form of high availiabity in which
clients ``stick to'' (i.e. have affinity with) at least one server (a
property which is often implicit in the distributed systems
literature~\cite{herlihy-art, cops, eiger, cac} but which requires
\textit{explicit} consideration in a client-server replicated database
context). This sticky availability is easily implementatable and
widely employed~\cite{cops, eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.
